A data-driven study of physiotherapy for cystic fibrosis: identifying sub-types of airway clearance and physical activity patterns using unsupervised clustering

### 2019-09-05 draft.tex
pushing all work for up to Sept 4th.

% to avoid loading the natbib package, add option nonatbib:
\usepackage[nonatbib]{neurips_2018}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}

\title{A data-driven study of cystic fibrosis treatment: identifying sub-types of physiotherapy style and physical activity using unsupervised clustering}

@@

\begin{abstract}
  Children and young people with cystic fibrosis (CYPwCF) are prescribed physiotherapy including daily airway clearance techniques (ACT), and being physically active as part of their treatment. Little is known about how CYPwCF follow this advice in their everyday lives. Physiotherapy has evolved over the decades without this detailed data. Such data is now being collected as part of Project [anonymous] which is the largest CF physiotherapy trial in the UK to date. The data comes from novel pressure sensors attached to existing ACT devices which remotely measure breathing patterns during normal treatments, and Fitbit activity trackers which monitor physical activity patterns. In close collaboration with clinical experts, we quantify and characterise this unique ACT and activity data by creating custom features of the waveforms. We identify sub-types with unsupervised learning, with k-means clustering achieving the best performance. We find 4 clusters of ACT data which are sensitive enough to track interventions in ACT routine, 4 clusters of steps data, and 5 clusters of heartrate data, which are clinically meaningful sub-types.
\end{abstract}

@@

Cystic Fibrosis (CF) is a chronic life limiting illness affecting approximately 1 in 2500 babies born globally. CF is a genetic disorder that mainly affects the lungs, where excessive thick sticky mucous leads to a cycle of infection, inflammation and lung damage leading to a deterioration in health. Despite recent progress CF remains progressive and incurable. In some cases, a lung-transplant is performed, but only half of the current CF population are expected to live to over 40 years old [*******UCL****** ref to this or similar].

To clear the lungs of excess mucous, daily physiotherapy airway clearance techniques (ACT) are prescribed. The physiotherapy exercises involve a series of forced exhalations into a handheld device  followed by a huff and cough to improve mucociliary clearance and expel mucous; an exercise which is performed repeatedly [*******UCL****** ref to guideline?]. For children with CF, the physiotherapy is burdensome, with some reporting that they dislike their physiotherapy more than they dislike the disease itself [*******UCL****** ref? This was from a survey that Jackie mentioned]. Physiotherapists also encourage children to have a physically active lifestyle [*******UCL****** ref to why physical activity is good]. Until recently adherence to physiotherapy recommendations was only assessed via questionnaire or self report which is known to be inaccurate [*******UCL****** ref]. The evolution of CF physiotherapy of the decades has not been informed by this important evidence.

Bespoke pressure sensors were developed to attach to existing ACT devices and remotely monitor ACT breathing patterns. One way to reduce the treatment burden of ACT may be by using gamification. To investigate if gamification can change physiotherapy adherence or patterns, bespoke games were developed which are driven by pressure changes measured with the sensor [ref 1: verbal description of games, ref 2: video demonstrating the point https://customers.microsoft.com/en-us/story/731791-fizzyo, removing any identifyin names in the bibliography]. Physical activity was also assessed using Fitbit activity trackers.

 145 CYPwCF participated in the study [NREC:18-LO-1038] to quantify ACT and physical activity patterns (via wearing of a Fitbit activity tracker), and to study the effects of gamification. This is the largest CF physiotherapy trial in the UK [*******UCL****** ref to meta-analysis of CF trials, or previously largest studies, or any previous data-driven studies?], collecting completely novel daily data about how children follow physiotherapy advice in their everyday lives. The broader goals of the study are to link physiotherapy behaviour patterns with clinical outcomes, in order to optimise and personalise treatment. In this paper we describe the challenge of i) quantifying ACT pressure sensor data and Fitbit activity data by creating custom features of the waveforms, and ii) discovering sub-types of ACTs and activity using clustering techniques. These daily sub-types can be tracked over time, as interventions such as gaming are introduced.   
\section{Methods}
\subsection{Trial design and analysis approach}
We used an interrupted time series design for the 18 month trial, so the effect on ACT patterns of introducing and removing an intervention can be assessed [[*******UCL****** ref to details of trial design,is this sufficient? NREC:18-LO-1038 ]. Interventions include introducing feedback to participants on their ACT (number of breaths) on a tablet application, introducing games, and later removing these interventions. The data presented here is from the first 8x months of the trial. Participants are: [*******UCL****** exact numbers] 65x males and 80x females ranging in age from 6-16 years old, from 3 London hospitals, and are voluntarily part of the study. All data is completely de-identified for analysis to researchers.

Our analysis seeks to find patterns in physiotherapy data. Once these patterns are known, they can be linked with clinical outcomes to determine if certain behaviour patterns are preferable. We do not impose pre-conceived labels such as “good” or “bad” on waveforms, thus we leave open the possibility that what is currently prescribed may not be optimal. As such, the data is unlabelled and we use unsupervised learning approach to achieve this. Unsupervised clustering has been used to find subtypes in physical activity data \cite{physical_activity_patterns_2017} and pressure data relating to respiratory disease [].

@@

\subsection{Collaboration, trust and interpretability}
Many applications of machine learning to medical data fail to explain how (if at all) the expertise of clinicians has been included [ref https://www.nature.com/articles/s41598-018-29523-2]. To address the issues that arise from a lack of collaboration [ref to stuff on collaboration and interpretability] our methodology involves close collaboration between data scientists and CF experts, making every data decision jointly. We use a novel approach of involving at least one clinical expert in engineering “stand-ups” every day over the course of building the data analysis pipeline (7 months). We only consider features that are meaningful to clinicians, and we balance traditional statistical evaluation metrics, with what is interpretable to clinicians.

\subsection{Feature engineering}
\subsubsection{Breath data description}
An ACT treatment commonly involves around 10 sets of 10 exhalation breaths into the ACT device. Each set loosens mucous, so that after the set the participant can huff and cough to expel it. The treatment may be performed once or twice per day; participants do not have identical prescriptions. The ACT device is fitted with a wireless pressure sensor sampling at 10 Hz, recording a pressure value (cm H2O) and its time stamp. Participants each used 1-3 of 4 different ACT devices compatible with the sensor; the Acapella, Aerobia, PariPEP and AstraPEP.  
\subsubsection{Processing breath data}
Several pre-processing steps were performed on the data which were a function of the custom sensor used: including assembling data packets and removing duplicate values. The sensor also introduced significant non-linear baseline drift. Because of the asymmetrical nature of the exhalation peaks and sparsity of the signal, we used a sparsity-based de-noising approach with an asymmetry factor, called BEADS \cite{Ning2014}, which is designed for use with chromatogram waveforms which have a similar form. 
% figure 1
In total, we recorded 5x,000 sensor readings. This includes data from every time the ACT sensor was turned on, including for testing and demonstration. Once we removed readings shorter than 15 seconds, treatments with no breath exhalations, and situations with sensor misuse/malfunction, we were left with 2x,000 ACT treatments for analysis.

\subsubsection{Featurising breath data}
% figure 2
The defining characteristics of ACT treatments were summarised in a range of features (variables): breath count, breath duration, breath amplitude, and treatment duration. In order to derive these features, our breath identification algorithm recognises breaths composed of multiple peaks, and having breath shapes which are characteristic of the 3 different ACT devices used in the study. In addition to mean breath amplitude, we use amplitude standard deviation as a feature, in order to capture inconsistency or disorderliness in physiotherapy. We also identified how many discernible sets were apparent in the treatment. These features are on a per-treatment level.

The aforementioned features capture the characteristics of existing treatments. However it is common for prescribed treatments to be skipped, and days with no treatments is an important signal itself. Several features were derived which capture missing data: treatment adherence to date, skipped days since last treatment, rolling adherence over 7 days. These features are on a per-day level, as participants are prescribed 1 or 2 treatments per day.

\subsubsection{Featurising physical activity data}
The Fitbit activity tracker (Fitbit AltaHR) provides each participant’s step count per minute. While real-time heartbeat data would allow for a detailed analysis of heartrate waveforms, the Fitbit only provides average heartrate, with a variable sampling frequency ranging from 3 to 12 values per minute. To achieve consistent frequency and comparable values, the heartrate values are averaged for each minute.

Mindful of these data limitations, we extract high-level patterns from the minute averaged waveforms at a day level. The heartrate and steps datasets are featurised using summary statistics per day using counts, frequency and interval values such as number of minutes in moderate to vigorous intensity physical activities (MVPA) (heartrate above 120 bpm) per day, step cadence (number of steps/minutes of wear) at day level, and number of hours with more than 500 steps at day level.

% figure 3

\subsubsection{Missing activity data}

To account for missing data during times that the Fitbit is not worn, the wear percent of the Fitbit is calculated from the heartrate values. This measure is used to select only days with a substantial portion of data for the day (80\% of wear time from 6 am to 6 pm) for analysis. The wear time is also used to normalise some features and to obtain step cadence for minutes where heartrate data is present.

@@

All features were verified by clinicians as being meaningful or potentially meaningful. We perform a correlation analysis on all features and removed features which were highly correlated. We use the Hopkins statistic [ref] to measure clustering tendency. Features with a Gaussian distribution are normalised to have mean =0 and standard deviation 1. In order to visualise the results, we use dimensionality reduction. We use the first two principle components from principal component analysis [ref], Uniform Manifold Approximation and Projection (UMAP), and t-Distributed Stochastic Neighbor Embedding (T-SNE). We identify outliers using xxx methods and remove less than 2\%, with these cases being recognised as device malfunctions or unusual human behaviour.

We employ the following clustering methods: k-means, Density-based spatial clustering of applications with noise (DBSCAN) and hierarchical clustering.

\section{Results}
\subsection{Clustering breath waveforms}
\subsubsection{Feature selection for clustering}
Some features were excluded from clustering, but provide interesting physiotherapy insights.  
The number of sets in a treatment were expected to be very influential in clustering, as a regular pattern of sets shows close adherence to physiotherapy advice. However we found that less than 20xx\% of treatments had discernible sets, so this feature could not be used

% table 1

\subsubsection{ACT clustering results}
The results of the four techniques used for clustering are show in table x. K-means was found to perform best because we liked the way..… K-means is a simple

% table 2
% table 3
% figure 4

@@

We have checked the clustering stability on different data partitions.  Random sampling of the data (80\% of data) and using k-means clustering provided very similar results comparing to clustering on the full dataset:  shape of the clusters, cluster centroids as well as Silhouette Coefficients and Calinski-Harabaz Index  (2200 v 1770 respectively) are comparable. 
% figure
We also have done validation of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters. Figure xx shows the same patient trajectory for the Clustering A and Clustering B. Clustering B uses breath amplitude features (mean and standard dev) as well as breathRollWeekAdherenceScore in addition to features that Clustering A uses. Notable that  once computer gaming is introduced to ACT in both cases A and B the same patient moves from cluster with “expected” breath counts per ACT to cluster characterized by much bigger breath counts. This validates that there is a change in patient’s behaviour that we’ve captured. Additional sdBreathAmplitude feature make results more explainable from clinical point of view as original cluster for the patient in Clustering B was characterized by centroid with the biggest sdBreathAmplitude value. 
% figure

@@

Clustering of physical activity data was used to extract daily patterns from the steps and heartrate waveforms that can be utilised to individually characterise the physical activity levels and patterns of individual patients. The different clusters should extract more complex characteristics than the ones which could be obtained by simply combining some of the extracted features.

@@

The following features were used for clustering steps: number of minutes with step count higher than 100 for at least 5 consecutive minutes, the mean of total step count per hour, the step cadence for the periods in MVPA and the number of hours in the day with 0 < steps count < 500. Features were selected as a result of feature selection, correlation and outlier analysis. Four clusters were found by Kmeans with the following characteristics:

\begin{itemize}
\item XS - this cluster contains waveforms with very little steps activity or much missing step data 
\item S - moderate step activity and consistent cadence throughout the day with very few periods of high step activity 
\item M - consistent moderate to high step rate across the day with period of sustained high step activity  
\item L - High step rate throughout the day with longer periods of sustained high step activity
\end{itemize}
%figure
%table

subsubsection{Heartrate clusters}

The number of minutes with heartrate higher than 120, coefficient of variation for heartrate values at a day level, number of time heartrate goes above 100 and 120 and the median average hourly heartrate were used as features for the heartrate clusters. Five clusters were found by Kmeans with the following chacteristics: 
\begin{itemize}
\item S - Small heartrate values and consistently below 120 
\item M - Consistent variations around 100 with very few spikes in heartrate values 
\item M1 - Waveform with high variability and periods with spikes above 120 and fluctuations above the 120 mark 
\item M2 - Consistent variations around 100 with spikes in values above 120 
\item L - Large values around and above 120 
\end{itemize}
%figure
%table

@@

The steps and heartrate clusters were validated with the clinical experts. Individual waveforms were manually inspected together with the cluster boundaries to ensure consistency across cluster interpretations and the actual cluster characteristics. Additionally, other features that were not included in the clustering analysis were used to validate the clusters and their interpretations. Certain expected conditions were checked against the cluster labels and plotted using the PCA graph such as Figure Xxx below.

@@

Project Fizzyo involves collecting novel data about how CF physiotherapy advice, ACTs and physical activity, is adhered to. In close collaboration with CF experts, we develop methods to interrogate this data. We derive clinically meaningful features from ACT and activity (heartrate and steps) waveforms. We use unsupervised clustering to find sub-types of this data.

Project Fizzyo is an ongoing study, with participation lasting 18 months. 7 months into the trial we do not draw any conclusions, as we do not have an equal number of days of data from patients. Instead, we present a framework for analysis, which will empower clinicians to interpret the impact of interventions such as gamification, and what type of physiotherapy leads to better clinical outcomes.

@@

\begin{figure}
  \centering
  \caption{Sample figure caption.}
  \includegraphics[]{sample_fig}
\end{figure}

\subsection{Tables}
\begin{table}
  \caption{Sample table title}
  \label{sample_table}
  \centering
  \begin{tabular}{lll}
    \toprule
    \cmidrule(r){1-2}
    Col1 & Col2 & Col3 \\
    \midrule
    a & 0 & z\\
    b & 1 & y \\
    c & 2 & x \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\appendix
\section{Appendix : Table of physical activity data features}

\begin{table}[h]
  \caption{Physical activity features}
  \label{fitbit_features}
  \centering
  \begin{tabular}{ |l|l|l| }
    \toprule
    \cmidrule(r){1-2}
    Feature Name & Feature description & Waveform extracted from \\
    \midrule
    userif & 1099 & 0.95 \\
    \bottomrule
    \end{tabular}
\end{table}

\section{Appendix : Cluster metrics for physical activity data}

>>>>

% to avoid loading the natbib package, add option nonatbib:
\usepackage[nonatbib]{neurips_2018}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\newcolumntype{n}{>{\arraybackslash}m{5.3cm}}
\newcolumntype{d}{>{\arraybackslash}m{8cm}}

\title{A data-driven study of physiotherapy for cystic fibrosis: identifying sub-types of airway clearance and physical activity patterns using unsupervised clustering}

@@

\begin{abstract}
  Children and young people with cystic fibrosis (CYPwCF) are prescribed physiotherapy including daily airway clearance techniques (ACT), and being physically active as part of their treatment. Little is known about how CYPwCF follow this advice in their everyday lives. Physiotherapy has evolved over the decades without the clarity of evidence that would accrue from detailed data. Such data are now being collected as part of Project [Anonymous] which is the largest CF physiotherapy trial in the UK to date. The data come firstly from novel pressure sensors attached to existing ACT devices which measure breathing patterns during normal treatments, and secondly from Fitbit activity trackers which monitor physical activity patterns. In close collaboration with clinical experts, we quantified and characterised this unique ACT and activity data by creating custom features from the waveforms. Using unsupervised learning techniques, We identified meaningful sub-types, with k-means clustering achieving the best performance. We found 4 clusters of ACT data which were sensitive enough to track interventions in ACT routines. Within physical activity data we identified 4 clusters related to daily footsteps  and 5 clusters related to heart rate data.
\end{abstract}

@@

Cystic Fibrosis (CF) is the most common life-limiting inherited disorder in the UK, affecting approximately 1 in 2500 babies born. It is a systemic genetic disorder that mainly affects the respiratory and digestive systems. In the lungs excessive thick sticky mucus can cause cycles of infection, inflammation and lung damage leading to a deterioration in health. Despite improvements in care, CF remains progressive and incurable. In some cases, lung-transplantation can prolong life, but only half of the current CF population are expected to live to over 40 years old \cite{Keogh2018}.

To promote clearance of mucus from the lungs, daily physiotherapy is prescribed \cite{Daniels2017}. Airway clearance techniques (ACT) may include a series of  exhalations against resistance into hand-held ACT devices, followed by forced expiration techniques ('huffing') and coughing to improve mucociliary clearance and expel mucous; an exercise which is performed repeatedly \cite{Villanuevaj4574}. ACT is burdensome, with some CYPwCF reporting that ACT is the most burdensome part of having the disease \cite{JamesLindAlliance}. Physiotherapists also encourage children to have a physically active lifestyle, which has been shown to reduce the rate of decline in lung function \cite{Williams2014}. Adherence to physiotherapy has traditionally been assessed via questionnaire or self report, both of which are known to be inaccurate. More objective measures of adherence  have not been widely available, and the evolution of CF physiotherapy has not been informed by this important evidence \cite{Modi2006}.

Bespoke pressure sensors were developed to attach to existing ACT devices in order to facilitate remote monitoring of ACT undertaken routinely at home. In addition, bespoke games were developed which could be driven by pressure changes from breathing through the sensor. This was in order to investigate if gamification might reduce the treatment burden of ACT or modify ACT adherence or patterns \cite{Jackson2019} \cite{[Anonymised]2019} . Physical activity patterns were also measured using Fitbit activity trackers, which generated granular detail on footsteps and heart rate changes during the day.

 145 CYPwCF participated in the study [NREC:18-LO-1038] to quantify routine ACTs (using bespoke pressure sensors) and physical activity patterns (from wearing a Fitbit activity tracker), and to study the effects of gamification. This is the largest CF physiotherapy trial in the UK [*******UCL****** ref to meta-analysis of CF trials, or previously largest studies, or any previous data-driven studies?], collecting completely novel daily data about how children follow physiotherapy advice in their everyday lives. The broader goals of the study are to link physiotherapy adherence patterns with clinical outcomes, in order to optimise and personalise treatment. In this paper we describe the challenge of i) quantifying ACT pressure sensor data and Fitbit activity data by creating custom features of the waveforms, and ii) discovering sub-types of ACT and physical activity patterns using clustering techniques. These daily sub-types can be tracked over time, as interventions such as gaming are introduced.

\section{Methods}

\subsection{Trial design and analysis approach}

The 18 month trial used an interrupted time series design, so that the effect of introducing and removing interventions such as gamification could be assessed [[*******UCL****** ref to details of trial design,is this sufficient? NREC:18-LO-1038 ]. Interventions included: introducing feedback to participants on their ACTs (number of breaths per treatment) via a computer tablet application, introducing games, and later removing these interventions. The data presented here is from the first 8x months of the trial. Participants were: [*******UCL****** exact numbers] 65x males and 80x females ranging in age from 6-16 years old, from 3 London hospitals, all of whom consented to participate in the study. All data were completely de-identified for analysis.

Our analysis seeks to find patterns in physiotherapy data. Once these patterns are known, they can be linked with clinical outcomes to determine if certain behaviour patterns are preferable. We do not impose pre-conceived labels such as “good” or “bad” on waveforms, thus we leave open the possibility that what is currently prescribed may not be optimal. As such, the data is unlabelled and we use unsupervised learning approach to achieve this. Unsupervised clustering is becoming an established technique for finding sub-types in physical activity data \cite{physical_activity_patterns_2017} and pressure data relating to respiratory disease [].

@@

\subsection{Collaboration, trust and interpretability}

Many applications of machine learning to medical data fail to explain how (if at all) the expertise of clinicians has been included \cite{Alaa2018}. To address the issues that arise from a lack of collaboration \cite{Vayena2018}, \cite{Challen231}, \cite{Char2018}, \cite{Ahmad2018} our methodology involved close collaboration between data scientists and CF experts, making every data decision jointly. We used a novel approach of involving at least one clinical expert in engineering “stand-ups” every day over the course of building the data analysis pipeline (7 months). We only considered features that were meaningful to clinicians, and we balanced traditional statistical evaluation metrics, with what was interpretable to clinicians.

\subsection{Feature engineering}

\subsubsection{Breath data description}

ACT treatments commonly involve around 10 sets of 10 exhalation breaths into an ACT device. Participants generally used 1 of 4 different ACT devices compatible with the pressure sensor; the Acapella, Aerobika, PariPEP or AstraPEP. Each device produced breaths with a slightly different overall shape, e.g. with small oscillations. Sometimes participants used different devices over different days. Each set helps to loosen mucus, and between sets, patients normally 'huff' and cough to expel it. The treatments are usually performed one or more times per day depending on how well the person is or how much mucus they normally produce; participants did not have identical prescriptions. For the study, the ACT device was fitted with a bespoke wireless pressure sensor sampling pressure at 10~Hz, recording pressure values in time-stamped millibars.

\subsubsection{Processing breath data}

Several pre-processing steps were performed on the data which were a function of the custom sensor used: including assembling data packets and removing duplicate values. The sensor also introduced significant non-linear baseline drift. Because of the asymmetrical nature of the exhalation peaks and sparsity of the signal, we used a sparsity-based de-noising approach with an asymmetry factor, called BEADS \cite{Ning2014}, which is designed for use with chromatogram waveforms which have a similar form. An example of pre-processing raw ACT data is shown in figure \ref{fig:ACTpreprocessing}.

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{ACTpreprocessing.png}
  \centering
   \caption{Left: raw ACT data in two packets, with baseline drift. Right: Data packets assembled into a treatment and baseline drift removed. This treatment shows 11 sets of 10 exhalations (the positive peaks), lasting 33 minutes.}
  \label{fig:ACTpreprocessing}
\end{figure}

In total, we recorded 5x,000 sensor readings. This included data from every time the ACT sensor was turned on, including for testing and demonstration. Once we removed readings shorter than 15 seconds, treatments with no breath exhalations, and situations with sensor misuse/malfunction, we were left with 2x,000 ACT treatments for analysis.

\subsubsection{Featurising breath data}

A typical ACT waveform which has been pre-processed, is shown in figure \ref{fig:individual_breaths}. The defining characteristics of ACT treatments were summarised in a range of features (variables) including: breath count, breath duration, breath amplitude, and treatment duration. In order to derive these features, our breath identification algorithm recognises not only simple breaths, but complex breaths composed of multiple peaks, and having a variety of shapes & oscillations which are characteristic of the 3 different ACT devices used in the study. In addition to mean breath amplitude, we use amplitude standard deviation as a feature, in order to capture inconsistency or disorderliness in physiotherapy. We also identified how many discernible sets were apparent in the treatment. These features are on a per-treatment level.

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{individual_breaths.png}
  \centering
   \caption{A typical, pre-processed ACT waveform. This treatment is 20 seconds long and contains 7 exhalation breaths.}
  \label{fig:individual_breaths}
\end{figure}

The aforementioned features capture the characteristics of existing treatments. However it is common for prescribed treatments to be skipped, and days with no treatments is an important signal itself. Several features were derived which captured missing data: treatment adherence to date, skipped days since last treatment, rolling adherence over 7 days. These features are on a per-day level, as participants are prescribed 1 or 2 treatments per day.

\subsubsection{Featurising physical activity data}
The Fitbit activity tracker (Fitbit AltaHR) provided each participant’s step count per minute. While real-time heartbeat data would allow for a detailed analysis of heart rate waveforms, the Fitbit only provides average heart rate, with a variable sampling frequency ranging from 3 to 12 values per minute. To achieve consistent frequency and comparable values, the heart rate values were averaged for each minute. An example of heart rate data is shown in figure \ref{fig:HR}.

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{HR.png}
  \centering
   \caption{One day showing heartrate data with a sustained period of heartrate in MVPA.}
  \label{fig:HR}
\end{figure}

Mindful of these data limitations, we extracted high-level patterns from the minute-averaged waveforms at a day level. The heart rate and steps data sets were featurised using summary statistics per day using counts, frequency and interval values, such as number of minutes in moderate to vigorous intensity physical activities (MVPA) (heart rate above 120 bpm) per day, step cadence (number of steps/minutes of wear) per day, and number of hours with greater than 500 steps per day.

To account for missing data during times that the Fitbit was not worn, the wear percent of the Fitbit was calculated from the valid heart rate values. This measure was used to select only days with a substantial portion of data for the day (80~\% of wear time from 6~am to 6~pm), resulting in a dataset containing 9760 days used for analysis. The wear time was also used to normalise some features and to obtain step cadence for minutes where heart rate data was present.

@@

All features were verified by clinicians as being meaningful or potentially meaningful. We perform a correlation analysis on all features and removed features which were highly correlated. We use the Hopkins statistic [ref] to measure clustering tendency. Features with a Gaussian distribution were normalised to have mean~=~0 and standard deviation~=~1. In order to visualise the results, we used dimensionality reduction. We used the first two principle components from principal component analysis [ref], Uniform Manifold Approximation and Projection (UMAP) [ref], and t-Distributed Stochastic Neighbor Embedding (T-SNE [ref]. We identified outliers using xxx (Olga) methods and removed less than xxx2~\%, with these cases being recognised as device malfunctions or abnormal behaviour.

We employed the following clustering methods: k-means, Density-based spatial clustering of applications with noise (DBSCAN) [ref?] and hierarchical clustering.

\section{Results}

\subsection{Clustering breath waveforms}

\subsubsection{Feature selection for clustering}

Some features were excluded from clustering, but provided interesting physiotherapy insights.  
The number of sets in a treatment were expected to be influential in clustering, as a regular pattern of sets would show close adherence to physiotherapy advice. However we found that less than 20xx\% (Tempest) of treatments had discernible sets, so this feature could not be used. The features used for clustering are shown in table X below, while the full list of features is shown in appendix X.

\subsubsection{Comparison of clustering techniques}
We compared the three clustering techniques (see appendix) and found the best performance with k-means.

\begin{figure}[!htb]
  \centering
  \includegraphics[width=\textwidth]{featureTables.png}
  \caption{To be replaced with a proper table.}
\end{figure}

\subsubsection{ACT K-means clustering results}
The purpose of clustering into sub-types is to summarise how a participant is performing their ACT. We chose k=4 clusters to maximise both silhouette value (average 0.18), and the interpretability of the results. If k is too low, the sub-types are not granular enough, if k is too high it becomes challenging to remember the characteristics of each sub-type.

The centroid values of the 4 clusters are shown in table X below. Based on these centroids, the four ACT sub-types, which we named alpha, beta, gamma and delta have the following interpretations:

\begin{itemize}
\item Alpha - Low adherence  ....
\item Beta - High inconsistency ....
\item Gamma -  ....
\item Delta -  ....
\end{itemize}

@@

The ACT sub-types should be sensitive enough to capture changes in ACT, particularly during interventions. Figure ~\ref{fig:act_res} shows that clusters are sensitive enough to track changes in a participants' adherence through different study interventions, which validates our approach.

\begin{figure}[!htb]
  \centering
  \includegraphics[]{fig_ACT_clust}
  \caption{the four ACT clusters overlaid with a single participant (black), showing that the participant’s cluster membership changes over time. Clustering results are visualised in 2-dimensions using PCA (left) and UMAP (right). The time points are before ACT feedback (number of registered breaths) is introduced (top), after ACT feedback is introduced (middle), and once computer gaming is introduced to ACT routine (bottom). Each point represents one treatment of one participant. }
  \label{fig:act_res}
\end{figure}

We also assessed the clustering stability on different data partitions. Random sampling 80~\% of the data and clustering using k-means provided very similar results with clustering on the full dataset: figure \ref{fig:ACTclusteringon80} shows that the shape of the clusters, cluster centroids as well as Silhouette Coefficients and Calinski-Harabaz Index (2200 v 1770 respectively) were comparable.

\begin{figure}[!htb]
  \centering
  \caption{K-means clustering results on full dataset (left) and subset of data (right).}
  \includegraphics[]{ACTclusteringOn80.png}
  \label{fig:ACTclusteringon80}
\end{figure}

*Not sure about this part. Maybe move to appendix?
We also assessed the validity of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters. Figure \ref{fig:ACTclusteringAB} shows the same patient trajectory for clustering with feature selection A and clustering with feature selection B, which includes two additional features: breath amplitude standard deviation, and rolling weekly adherence score for breath count. It is notable that once computer gaming was introduced to ACT in both cases A and B the same patient moved from cluster with “expected” breath counts per ACT to a cluster characterized by much higher breath counts. This validates that there was a change over time in the patient’s ACT pattern that could be captured. The additional sdBreathAmplitude feature make results more explainable from a clinical point of view, as the original cluster for the patient in Clustering B was characterized by a centroid with the highest sdBreathAmplitude value.

\begin{figure}[!htb]
  \centering
  \caption{Patient Trajectory on set A of features (without breath amplitude) and set B of features.}
  \includegraphics[]{ACTclusteringAB.png}
  \label{fig:ACTclusteringAB}
\end{figure}

@@

The purpose of clustering into sub-types is to summarise each participant's physical activity data. Clustering of physical activity data was used to extract daily patterns from the steps and heart rate waveforms and characterise the physical activity levels and patterns of individual patients using a dataset containing 9760 individual days. The different clusters should extract more complex characteristics than the ones which could be obtained by simply combining some of the extracted features.

@@

The following features were used for clustering steps: number of minutes with step count higher than 100 for at least 5 consecutive minutes, the mean of total step count per hour, the step cadence for the periods in MVPA and the number of hours in the day with 0~< ~steps~count~<~500. Features were selected as a result of feature selection, correlation and outlier analysis. Three sub-types were desired for the steps clusters covering small, medium and large amounts of steps activity throughout the day. Four clusters (as shown in figure \ref{fig:stepsClusters}) were chosen in this case to cover these three sub-types as well as a sub-type for the instances with missing data (XS cluster).

\begin{figure}[htb]
  \centering
  \caption{Steps clusters displayed on the first two principle components.}
  \includegraphics[]{stepsClusters.png}
  \label{fig:stepsClusters}
\end{figure}

\begin{table}[H]
  \caption{Steps Clusters Interpretation}
  \label{steps_inter}
  \centering
  \begin{tabular}{ c|l }
    \toprule
    \cmidrule(r){1-2}
    Cluster name & Cluster Interpretation \\
    \hline
    XS & Waveforms with very little steps activity or high amount of missing data \\
    \hline
    S & Moderate step activity, consistent cadence with few periods of high activity\\
    \hline
    M & Consistent moderate to high step rate with a period of sustained high step activity \\
    \hline
    L & High step rate with longer periods of sustained high step activity \\
    \bottomrule
    \end{tabular}
\end{table}

\subsubsection{Heart rate clusters}

The following were used as features for heart rate clustering: the number of minutes with heart rate higher than 120, coefficient of variation for heart rate values per day, number of time heart rate exceeds 100 and 120, and the median average hourly heart rate. Experiments were carried out with 3, 4 and 5 clusters and the choice of k was based on the cluster interpretability results. The sub-types found when k=5 were the best at identifying more complex patterns and grouping similar waveforms together. 
The five clusters found by K-means (as shown in figure \ref{fig:HRclusters} have the following interpretation:

\begin{figure}[!htb]
  \centering
  \includegraphics[]{HRclusters.png}
  \caption{PCA of representation of the five heart rate clusters. }
  \label{fig:HRclusters}
\end{figure}

\begin{table}[H]
  \caption{Heartrate Clusters Interpretation}
  \label{hr_inter}
  \centering
  \begin{tabular}{ c|l }
    \toprule
    \cmidrule(r){1-2}
    Cluster name & Cluster Interpretation \\
    \hline
    S & Small heart rate values and heart rate consistently below 120 \\
    \hline
    SM & Consistent variations around 100 with very few spikes in heart rate values \\
    \hline
    M1 &  Waveform with high variability and periods with spikes above 120 \\
    \hline
    M2 & Consistent variations around 100 with spikes in values above 120 \\
    \hline
    L &  Large values around and above 120 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

The steps and heart rate clusters were validated with clinical experts. The clusters were found to be meaningful and useful ways to categorise the waveforms. Individual waveforms were manually inspected together with the cluster boundaries to ensure consistency across cluster interpretations and the actual cluster characteristics. Additionally, other features that were not included in the clustering analysis were used to validate the clusters and their interpretations. Certain expected conditions were checked against the cluster labels and plotted using the PCA graph such as figure \ref{fig:FitbitClusterValida} below. Cluster boundaries for both steps and heartrate clusters were also utilised to validate the cluster properties and interpretation, these can be found in the Appendix.

\begin{figure}[!htb]
  \centering
  \includegraphics[]{FitbitClusterValidation.png}
  \caption{: Crosses are days with daily cadence above 30 which are all part of the L cluster.}
  \label{fig:FitbitClusterValidation}
\end{figure}

@@

Project [Anonymised] involves collecting novel data about how CF physiotherapy advice on ACTs and physical activity is adhered to in everyday life, information which has never been available to physiotherapists. In close collaboration with CF experts, we developed methods to interrogate and make sense of this data. Project [Anonymised] is an ongoing study, with participation lasting 18 months. 8 months into the trial we do not draw any conclusions, as we do not have an equal number of days of data from patients recruited over 8 months. Instead, we present a framework for analysis, which will empower clinicians to understand how physiotherapy advice is actually adhered to, interpret the impact of interventions such as gamification when the trial is complete., and eventually understand what type of physiotherapy leads to better clinical outcomes. Our framework involves processing the ACT and Fitbit data, deriving clinically meaningful features from it which characterise this novel data, and then using unsupervised clustering to find sub-types of this data. The clusters provide a useful way of tracking participants' trajectories across sub-types over time.

@@

\subsection{Tables}
\begin{table}
  \caption{Sample table title}
  \label{sample_table}
  \centering
  \begin{tabular}{lll}
    \toprule
    \cmidrule(r){1-2}
    Col1 & Col2 & Col3 \\
    \hline
    a & 0 & z\\
    b & 1 & y \\
    c & 2 & x \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\appendix

\section{Appendix : Table of physical activity data features}
